A Stopped Negative Binomial Distribution

### 2016-06-15 stat-and-prob-letters/snb-paper.tex
More changes based on Dan’s comments.

\begin{table}[t!]
\caption{Characteristics of the Stopped Negative Binomial Distribution}
\label{tab:snb}
\begin{center}
\begin{tabular}{|c|l|} \hline
Parameters & $p$ the success probability ($0\leq p \leq 1$; $q = 1-p$) \\
           & $s$ the number of successes before stopping ($s=1, 2, ...$)\\
           & $t$ the number of failures before stopping ($t=1, 2, ...$)\\ \hline
Support & min$(s,t) \leq k \leq s+t-1$  \\ \hline
$\mathbb{P}[Y=k]$ & ${k-1 \choose s-1} p^s (1-p)^{k-s} + {k-1 \choose t-1} (1-p)^s p^{k-t}$\\ \hline
$\mathbb{P}[Y \leq k]$ & $2 - \mathcal{I}_{q}(k+1, s) - \mathcal{I}_{p}(k+1, t)$\\ 
    & where $\mathcal{I}$ is the regularized incomplete beta function.\\ \hline
Mean & $\frac{s}{p} \mathcal{I}_p(s,t) + \frac{p^{s-1} q^{t-1}}{B(s,t)} +
  \frac{t}{q} \mathcal{I}_q(t,s) + \frac{p^{t-1} q^{s-1}}{B(s,t)}$\\ 
  & where $B$ is the beta function \\ \hline
MGF & $\left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s) $\\ \hline
Posterior & 
${k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)}+
{k-1 \choose t-1} \frac{B\left(\alpha + k-t, t+\beta\right)}{B(\alpha,\beta)}$\\
& for Beta($\alpha$, $\beta$) prior. \\ \hline
\end{tabular}
\end{center}
\end{table}

In this work we derive the distribution of the number of enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). Some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
The rest of this paper derives these results
and explores properties of the distribution.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the Binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the posterior distribution using a Beta prior distribution
for $p$. Section 6 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

@@

There is a close connection between the tail probabilities of the SNB and the 
binomial distributions.
The probability of reaching the success endpoint in an 
SNB($p$, $s$, $t$) random variable is 
equal to the probability of at least $s$ successes in a Binomial distributed 
random variable with size $s+t-1$ and success probability $p$. 
Likewise, the probability of reaching the failure endpoint is equal
to the probability of at most $s-1$ successes in a Binomial distribution. 
That is, the probability that the number of successes is at least $s$ in the 
Binomial model is the same that the trial succeeds (reaches $s$) in the SNB 
model.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [\text{trial succeeds}].
\end{equation}
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{equation*}
\mathbb{P}[B \geq s] = 1 - \mathcal{I}_{1-p}.
\end{equation*}
%\begin{align*}
%\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \mathcal{I}_{1-p}.
%\end{align*}
The corresponding SNB probability is
\begin{equation*}
\mathbb{P} [\text{trial succeeds}] 
  = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{equation*}
Let $i=k-s$. Using the fact that
\begin{equation*}
{i+s-1 \choose s-1} = {i+s-1 \choose i}
\end{equation*}
the last summation can be rewritten as
\begin{align}
\mathbb{P} [\text{trial succeeds}] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

@@

\section{The Compound Distribution}

Let us consider a Bayesian analysis with a Beta prior distribution on $p$.
\begin{prop} \label{prop:bayesian}
The compound distribution of the Stopped Negative Binomial 
distribution with a Beta($\alpha$, $\beta$) prior is
\begin{align} \label{eqn:posterior}
\mathbb{P} [Y = k | s, t, \alpha, \beta ] &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $k$ is in the range
$\min(s,t) \leq k \leq s+t-1$. When 
this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose t-1} p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose t-1} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp.
\end{align*}
The result in (\ref{eqn:posterior}) follows by the definition of the 
Beta function.
\end{proof}

@@

\begin{figure}[t!]
\centering
\subfloat[The expected sample size.]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
\hfill
\subfloat[The variance of the sample size.]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
\caption{
The expected size and variance of the compound SNB varying the uncertainty 
in the prior distribution of $p$. A larger value of $c$
corresponds to a Beta prior with smaller variance.
}
\label{fig:bayesian-sample-size}
\end{figure}

Fig.~\ref{fig:bayesian-sample-size} shows how the expected sample size
and variance of the compound SNB. As the certainty increases the 
compound distribution sample values converge to corresponding expected 
values with 
$p = 0.045$. It may be noted that convergence is relatively slow. This
is because when $c$ is small, the prior is heavily skewed. As $c$ increases
this effect is minimized.

>>>>

\begin{table}[t!]
\caption{Characteristics of the Stopped Negative Binomial Distribution}
\label{tab:snb}
\begin{center}
\begin{tabular}{|c|l|} \hline
Parameters & $p$ the success probability ($0\leq p \leq 1$; $q = 1-p$) \\
           & $s$ the number of successes before stopping ($s=1, 2, ...$)\\
           & $t$ the number of failures before stopping ($t=1, 2, ...$)\\ \hline
Support & min$(s,t) \leq k \leq s+t-1$  \\ \hline
$\mathbb{P}[Y=k]$ & ${k-1 \choose s-1} p^s (1-p)^{k-s} + {k-1 \choose t-1} (1-p)^s p^{k-t}$\\ \hline
$\mathbb{P}[Y \leq k]$ & $2 - \mathcal{I}_{q}(k+1, s) - \mathcal{I}_{p}(k+1, t)$\\ 
    & where $\mathcal{I}$ is the regularized incomplete beta function.\\ \hline
Mean & $\frac{s}{p} \mathcal{I}_p(s,t) + \frac{p^{s-1} q^{t-1}}{B(s,t)} +
  \frac{t}{q} \mathcal{I}_q(t,s) + \frac{p^{t-1} q^{s-1}}{B(s,t)}$\\ 
  & where $B$ is the beta function \\ \hline
MGF & $\left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s) $\\ \hline
Compound Distribution & 
${k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)}+
{k-1 \choose t-1} \frac{B\left(\alpha + k-t, t+\beta\right)}{B(\alpha,\beta)}$\\
& for Beta($\alpha$, $\beta$). \\ \hline
\end{tabular}
\end{center}
\end{table}

In this work we derive the distribution of the number of enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). Some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
The rest of this paper derives these results
and explores properties of the distribution.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the Binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the compound distribution using a Beta distribution
for $p$.

@@

There is a close connection between the tail probabilities of the SNB and the 
binomial distributions.
The probability of reaching the success endpoint in an 
SNB($p$, $s$, $t$) random variable is 
equal to the probability of at least $s$ successes in a Binomial distributed 
random variable with size $s+t-1$ and success probability $p$. 
Likewise, the probability of reaching the failure endpoint is equal
to the probability of at most $s-1$ successes in a Binomial distribution. 
That is, the probability that the number of successes is at least $s$ in the 
Binomial model is the same that the treatment succeeds (reaches $s$) in the SNB 
model.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [\text{treatment succeeds}].
\end{equation}
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{equation*}
\mathbb{P}[B \geq s] = 1 - \mathcal{I}_{1-p}.
\end{equation*}
%\begin{align*}
%\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \mathcal{I}_{1-p}.
%\end{align*}
The corresponding SNB probability is
\begin{equation*}
\mathbb{P} [\text{treatment succeeds}] 
  = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{equation*}
Let $i=k-s$. Using the fact that
\begin{equation*}
{i+s-1 \choose s-1} = {i+s-1 \choose i}
\end{equation*}
the last summation can be rewritten as
\begin{align}
\mathbb{P} [\text{treatment succeeds}] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

@@

\section{The Compound Probability Distribution}

Let us consider a Bayesian analysis with a Beta distribution on $p$.
\begin{prop} \label{prop:bayesian}
The compound distribution of the Stopped Negative Binomial 
distribution when $p$ is distributed Beta($\alpha$, $\beta$) is
\begin{align} \label{eqn:posterior}
\mathbb{P} [Y = k | s, t, \alpha, \beta ] &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $k$ is in the range
$\min(s,t) \leq k \leq s+t-1$. When 
this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose t-1} p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose t-1} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp.
\end{align*}
The result in (\ref{eqn:posterior}) follows by the definition of the 
Beta function.
\end{proof}

@@

\begin{figure}[t!]
\centering
\subfloat[The compound distribution's expected sample size.]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
\hfill
\subfloat[The compound distribution's variance.]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
\caption{
The expected size and variance of the compound SNB with $s=2$, $t=12$
with $\mathbb{E}\ p = 0.045$ varying the uncertainty 
in the distribution of $p$. A larger value of $c$
corresponds to a Beta with smaller variance.
}
\label{fig:bayesian-sample-size}
\end{figure}

Fig.~\ref{fig:bayesian-sample-size} shows how the expected sample size
and variance of the compound SNB. As the certainty increases the 
compound distribution sample values converge to corresponding expected 
values with 
$p = 0.045$. It may be noted that the compound distribution's mean
approaches the deterministic limit slowly. This
is because when $c$ is small, the distribution of $p$ is heavily skewed. 
As $c$ increases this effect is minimized.
